---
title: "ORFquant: splice-aware quantification of translation"
author: "Lorenzo Calviello"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{ORFquant: splice-aware quantification of translation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup_vig, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

*ORFquant* detects and quantifies ORF translation using Ribo-seq [1] data,
taking into account the presence of multiple transcripts expressed by each
gene [2]. It selects the transcripts supported by the data, finds ORFs on them
*de novo*, quantifies their translation and annotates each ORF, in transcript
and genomic space.

This vignette runs the three steps of an analysis on the small example data
included in the package:

1. `prepare_annotation_files()`: parse the annotation and link it to the genome
   sequence, once per annotation;
2. `prepare_for_ORFquant()`: compute P-site positions and junction reads from a
   BAM file, once per sample;
3. `run_ORFquant()`: find and quantify ORFs.

```{r load_lib, message = FALSE}
library(ORFquant)
```


## Example data

The example data cover 7 genes on human chr22 (*TXNRD2*, *RANBP1*, *PPIL2*,
*TOP3B*, *DDT*, *TCN2* and *PISD*):

- `chr22_example.gtf.gz`: their annotation, from GENCODE 47 [3];
- `chr22_example.fa.gz`: the sequence of chr22 (GRCh38), with every base more
  than 1 kb from these genes replaced by N. It is compressed with `bgzip` and
  indexed (`.fai` and `.gzi` files);
- `chr22_example.bam`: the Ribo-seq reads aligned to these genes, from a sample
  of human aortic endothelial cells (SRA run SRR15513199, GEO sample
  GSM5527724) [4], aligned with STAR;
- `chr22_example_cutoffs.tsv`: the P-site offset for each read length in this
  sample.

`inst/scripts/make_example_data.R` in the package source describes how they
were made.

```{r example_data}
extdata <- system.file("extdata", package = "ORFquant")
list.files(extdata)
```

All output files go to a temporary directory:

```{r out_dir}
out_dir <- file.path(tempdir(), "ORFquant_example")
dir.create(out_dir)
```


## Prepare the annotation

`prepare_annotation_files()` reads a GTF file and the genome sequence, and
writes a *TxDb* and an `_Rannot` file to `annotation_directory`. The `_Rannot`
file holds the annotated transcript and genomic features (exons, CDS, UTRs,
introns, start and stop codons, ...) that the other steps use. The genome
sequence is a FASTA file indexed with `samtools faidx` (`genome_seq`); the
`_Rannot` file refers to it by its path.

```{r prepare_annotation, results = "hide", message = FALSE, warning = FALSE}
prepare_annotation_files(annotation_directory = out_dir,
                         gtf_file = file.path(extdata, "chr22_example.gtf.gz"),
                         genome_seq = file.path(extdata, "chr22_example.fa.gz"),
                         forge_BSgenome = FALSE)
annotation_file <- file.path(out_dir, "chr22_example.gtf.gz_Rannot")
```

(*txdbmaker* warns that it ignores the phase of the `stop_codon` lines and that
the *TxDb* has no genome version; both are harmless.)

`load_annotation()` loads it into two global variables: `GTF_annotation`, a
list of the annotated features, and `genome_seq`, the genome sequence. The
other functions load the annotation themselves: run each analysis in a new R
session, and don't load two annotations in one session.

```{r load_annotation}
load_annotation(annotation_file)
names(GTF_annotation)
```

For example, for the *TCN2* transcript ENST00000698268.1, its exons, its CDS
in transcript coordinates, its gene and transcript ids, names and biotypes, and
its CDS sequence, translated with the genetic code of its chromosome:

```{r annotation_features}
GTF_annotation$exons_txs[["ENST00000698268.1"]]
GTF_annotation$cds_txs_coords["ENST00000698268.1"]
GTF_annotation$trann[GTF_annotation$trann$transcript_id == "ENST00000698268.1", ]
GTF_annotation$genetic_codes
cds <- GTF_annotation$cds_txs[["ENST00000698268.1"]]
Biostrings::translate(unlist(getSeq(genome_seq, cds)))
```


## Prepare the P-sites

`prepare_for_ORFquant()` needs the Ribo-seq reads, aligned to the genome with
a spliced aligner, sorted by coordinate and indexed, and a table of the
P-site offset (`cutoff`, the distance of the P-site from the read's 5' end)
for each read length. The `compartment` column is `nucl` for all chromosomes
that are not circular, or the name of a circular one, such as `chrM`.

```{r cutoffs}
read.delim(file.path(extdata, "chr22_example_cutoffs.tsv"))
```

It writes the P-site positions, of all reads and of uniquely mapping reads
(mapping quality above 50), and the number of reads on each annotated
exon-exon junction to `<dest_name>_for_ORFquant`:

```{r prepare_psites, results = "hide"}
prepare_for_ORFquant(annotation_file = annotation_file,
                     bam_file = file.path(extdata, "chr22_example.bam"),
                     path_to_rl_cutoff_file = file.path(extdata, "chr22_example_cutoffs.tsv"),
                     dest_name = file.path(out_dir, "SRR15513199"))
for_ORFquant_file <- file.path(out_dir, "SRR15513199_for_ORFquant")
```

The [Ribo-seQC](https://github.com/lcalviell/Ribo-seQC) package [5] also
writes such a file, together with statistics on the Ribo-seq data.

```{r psites}
for_ORFquant <- get(load(for_ORFquant_file))
names(for_ORFquant)
for_ORFquant$P_sites_all
for_ORFquant$junctions[for_ORFquant$junctions$reads > 0]
```


## Run ORFquant

`run_ORFquant()` analyses each genomic region with P-sites, here with one
core. `gene_name`, `gene_id` or `genomic_region` restrict the analysis to
part of the annotation.

```{r run_orfquant, results = "hide"}
ORFquant_results <- run_ORFquant(for_ORFquant_file = for_ORFquant_file,
                                 annotation_file = annotation_file,
                                 n_cores = 1,
                                 prefix = file.path(out_dir, "SRR15513199"))
```

It writes the results to `<prefix>_final_ORFquant_results` (load them with
`get(load())`), the detected ORFs and the selected transcripts to
`<prefix>_Detected_ORFs.gtf`, the `ORFs_tx` table below, one row per ORF, to
`<prefix>_Detected_ORFs.tsv`, and the ORFs' protein sequences to
`<prefix>_Protein_sequences.fasta`:

```{r output_files}
list.files(out_dir, pattern = "^SRR15513199")
```

The results are a list:

```{r results}
names(ORFquant_results)
```

`ORFs_tx` has one range per ORF, in transcript coordinates, with its P-sites,
p-values, annotation and `ORFs_pM` (P-sites per ORF length, scaled to sum to a
million over all ORFs, akin to TPM). `ORF_category_Tx` compares the ORF with
the CDS annotated on the same transcript (for example `ORF_annotated`, `uORF`,
`dORF` or `N_truncation`; `novel` on transcripts without an annotated CDS), and
`ORF_category_Gen` with the longest CDS annotated in the ORF's gene, in genomic
coordinates (for example `exact_start_stop` or `Alt3_start`; `novel` and
`novel_*` for ORFs that overlap no annotated CDS). `?ORFquant_output` describes
every column and category:

```{r orfs_tx}
orfs <- ORFquant_results$ORFs_tx
data.frame(gene = orfs$gene_name, category_tx = orfs$ORF_category_Tx,
           category_gen = orfs$ORF_category_Gen, P_sites = round(orfs$P_sites, 1),
           pval = signif(orfs$pval, 2), ORFs_pM = round(orfs$ORFs_pM),
           row.names = NULL)
```

On *TCN2*, ORFquant finds the annotated ORF of ENST00000698268.1 and a short
ORF on ENST00000450638.5. That transcript has an extra exon, and its annotated
CDS starts after it. The short ORF starts at the same codon as the annotated
ORF of ENST00000698268.1 but ends in the extra exon: it is a `uORF` on its
transcript, and `Alt5_stop` in genomic space. `ORFs_gen` has the ORFs' genomic
coordinates, one range per exon:

```{r orfs_gen}
tcn2_orfs <- orfs$ORF_id_tr[orfs$gene_name == "TCN2"]
ORFquant_results$ORFs_gen[names(ORFquant_results$ORFs_gen) %in% tcn2_orfs]
```

`selected_txs` lists the transcripts selected for quantification:

```{r selected_txs}
ORFquant_results$selected_txs
```

`?run_ORFquant` and `?ORFquant` describe every element of the results, and
`?ORFquant_output` the columns of `ORFs_tx`.


## Plots and report

`plot_ORFquant_results()` summarises the results in plots, for example of the
number of ORFs in each category, written as PDF files to `<results>_plots`. It
saves them in an `_ORFquant_plots_RData` file too. (This example has so few
ORFs that some groups in these plots have fewer than two; *ggplot2* can't draw
their distributions and warns about it.)

```{r plot_results, results = "hide", message = FALSE, warning = FALSE}
results_file <- file.path(out_dir, "SRR15513199_final_ORFquant_results")
plot_ORFquant_results(for_ORFquant_file = for_ORFquant_file,
                      ORFquant_output_file = results_file,
                      annotation_file = annotation_file)
plots_file <- file.path(out_dir, "SRR15513199_final_ORFquant_results_plots",
                        "SRR15513199_ORFquant_plots_RData")
```

```{r show_plot, fig.width = 14, fig.height = 4, out.width = "100%"}
ORFquant_plots <- get(load(plots_file))
names(ORFquant_plots)
ORFquant_plots[["ORFs_found"]]
```

This plot counts the ORFs by `ORF_category_Tx_compatible`, their category on a
transcript compatible with their structure (`compatible_tx`, a protein-coding
one if possible), so its counts differ from those of `ORF_category_Tx` above.

`create_ORFquant_html_report()` puts these plots into an HTML report, for one
or more samples. It needs [pandoc](https://pandoc.org):

```{r report, eval = FALSE}
create_ORFquant_html_report(input_files = plots_file,
                            input_sample_names = "SRR15513199",
                            output_file = file.path(out_dir, "SRR15513199_ORFquant_report.html"))
```

`plot_orfquant_locus()` plots the P-sites, junction reads, transcripts and ORFs
of one gene to a PDF file; it needs the *Gviz*, *lemon*, *dplyr* and
*GenomeInfoDb* packages:

```{r locus_plot, eval = FALSE}
plot_orfquant_locus(locus = orfs$gene_id[orfs$gene_name == "TCN2"][1],
                    orfquant_results = ORFquant_results,
                    bam_files = NULL,
                    plotfile = file.path(out_dir, "TCN2.pdf"))
```


## Running ORFquant on your own data

Your data need:

- a GTF file with `exon` and `CDS` lines, each with `transcript_id` and
  `gene_id`;
- the genome sequence as a FASTA file indexed with `samtools faidx`, with the
  same chromosome names as the GTF and the BAM file;
- the Ribo-seq reads aligned to the genome with a spliced aligner such as STAR,
  sorted by coordinate and indexed with `samtools index`;
- the P-site offsets for each read length, as above.

A whole transcriptome takes much longer than this example: for the whole
sample SRR15513199 and GENCODE 47, `run_ORFquant()` took 55 minutes
with `n_cores = 36` and used about 63 GB of memory. Run it on a compute
cluster; `n_cores` above 1 uses forked processes, on Linux or macOS. Run each
step in a new R session, for example with `Rscript`.

The script `run_orfquant.R`, installed with ORFquant, runs the three steps in
one R process from the command line, with their default parameters, and lists
the files it writes:

```sh
Rscript $(Rscript -e 'cat(system.file("scripts", "run_orfquant.R", package = "ORFquant"))') \
  --gtf genes.gtf --fasta genome.fa --bam sample.bam \
  --offsets sample_cutoffs.tsv --outdir results --cores 4
```

With `--annotation results/annotation/genes.gtf_Rannot` instead of `--gtf` and
`--fasta`, it reuses the annotation for other samples. `--gene-names` and
`--gene-ids` restrict the analysis to some genes, and `--help` lists all
options.


## References

1. Ingolia NT, Ghaemmaghami S, Newman JRS, Weissman JS (2009). Genome-wide
   analysis in vivo of translation with nucleotide resolution using ribosome
   profiling. *Science* 324, 218–223.
2. Calviello L, Hirsekorn A, Ohler U (2020). Quantification of translation
   uncovers the functions of the alternative transcriptome. *Nature
   Structural & Molecular Biology* 27, 717–725.
3. Mudge JM et al. (2025). GENCODE 2025: reference gene annotation for human
   and mouse. *Nucleic Acids Research* 53, D966–D975.
4. Chothani SP et al. (2022). A high-resolution map of human RNA translation.
   *Molecular Cell* 82, 2885–2899.
5. Calviello L, Sydow D, Harnett D, Ohler U (2019). Ribo-seQC: comprehensive
   analysis of cytoplasmic and organellar ribosome profiling data. *bioRxiv*
   601468.


## Session info

```{r session_info}
sessionInfo()
```
